The Impact of Model Assumptions on RES Drought Analysis

### 2024-11-14 Article/main.tex
Add the rest of comments to article

@@

Detailed data analysis, supported by access to additional data from EirGrid, revealed that the installed capacity alone does not accurately reflect the true capacity used by the grid. Instead, the Maximum Export Capacity (MEC) values from the spreadsheet are used as a basis. To adjust for actual grid conditions, these MEC values are further multiplied by a factor of 1.4, which provides a more realistic representation of the PV generation potential.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis dataset~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used in the models to generate the Atlite and C3S-Energy datasets.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and actual trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: global horizontal irradiance (referred to here as $ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles found in Germany and replicated in European countries~\cite{saintdrenan2018solar}.

@@

Given that PV capacity in Ireland remains relatively low in 2023, and in order to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These figures are based on the target energy capacity outlined in the roadmap published by EirGrid~\cite{eirgrid2023future}. In this scenario, new PV time series are generated for both the Atlite and C3S-E G models, while the wind CF time series remain the same, as significant changes in wind farm distribution are not expected. From there, the results of these new time series are combined using updated weights which reflect a more balanced scenario (9 GW for wind and 8 GW for PV).

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{droughts_methodology}
	\caption{Wind time series of CF (red) and its 24-hour moving average (green) from the 7th to the 15th of July 2021. CF threshold (dashed black). The grey bar shows the period identified as a wind drought under our definition}
	\label{fig:find_res_droughts}
\end{figure}

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve} a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. A wind CF time series for Ireland for all wind turbine power curve options and all smoothings is generated. The performance of each wind turbine power curve is then, calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The smoothing of the wind turbine power curve represents losses due to the wake effect, which is important when modelling wind energy on larger scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind data from EirGrid.

@@

The effect of the differences in the power curves are also visible in Fig.~\ref{fig:wind_verification_contour} where the Atlite model shows good agreement with the EirGrid data. In contrast, the two C3S-E datasets tend to overestimate the EirGrid CF. This is confirmed by Table~\ref{tab:wind_skill_scores}, which shows that Atlite performs better than the C3S-E datasets for all three skill scores.

@@

The C3S-E datasets calculate PV generation multiple times using the same model with varying azimuth and tilt angles. The results are then aggregated based on a statistical distribution of module angles of Germany and then adapted to other European countries. Similarly to the wind CF verification, the Atlite model allows the user to select a given PV panel characteristics. In this study we have chosen to look at the three PV panels offered by in the Atlite model. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Figure \ref{fig:solar_verification_contour} shows that the three datasets have a similar tendency of overestimating the CF compared to EirGrid. The skill scores presented in Table~\ref{tab:pv_skill_scores} indicate that C3S-E G performs best overall, with the lowest RMSE and a high correlation coefficient, suggesting a closer match to observed data. Atlite shows a slightly lower correlation and a higher RMSE, indicating larger deviations. All models show a slight positive bias, with Atlite exhibiting the most overestimation. These results reflect the broader challenges in accurately modeling PV, where variability is hard to capture. Nonetheless, C3S-E G’s lower RMSE suggests it may be more reliable for PV CF modeling in this context.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity. Due to uncertainties in PV generation data during the first few months after each farm is connected, it is difficult to accurately determine the effective capacity. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events. While the results are reasonably aligned in some ranges, the inherent challenges in PV data highlight the limitations of this verification.

@@

The median, first, and third quartiles for the Atlite dataset are consistently greater than or equal to those of the other two datasets, regardless of the duration range or type of renewable energy considered. This difference arises from the wind turbine power curve model used in the C3S-E datasets, which tends to overestimate the wind CF (Fig.~\ref{fig:wind_verification_contour}). As a result, the overall number of RES droughts is underestimated in the C3S-E datasets compared to Atlite.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves significantly different than the two C3S-E datasets. The Atlite results follow the ones for wind with a log-linear increase, but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold. Atlite appears to hover slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

@@

Similar to previous results, the existing installed capacity wind-PV combination (Fig.\ref{fig:res_droughts_seasonality}c) shows patterns comparable to the wind panel (Fig.\ref{fig:res_droughts_seasonality}a). However, in this panel, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of PV generation during the summer months in the existing scenario, even though it constitutes only 11\% of total capacity. Since the number of RES drought hours for PV in summer is near zero, this small contribution has a noticeable impact on reducing overall drought hours. In the projected scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. The balanced mix of wind and PV in this scenario reduces the seasonal signal, slightly increasing the percentage of RES drought hours in winter months while significantly decreasing it in summer. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for Atlite, from 8\% to 3\% for C3S-E G, and from 9\% to 4\% for C3S-E N.

@@

>>>>

\textcolor{magenta}{Some things to keep in mind when reading from Damian's comments:
	1. The whole current/projected and existing/balanced thing. Pick one and stick to it, including the figures (also fix the a/c level in fig 10)
	2. See if you are happy with the section titles throughout the document
	3. Check if when it is EirGrid/SONI it is clear, both in the text and references
	4. Is the Ireland / ROI / NI distinction clear enough? I did not see the line where we say that we call ROI+NI=Ireland (may have missed it) but I am alright with how it is written now either way}

@@

Detailed data analysis, supported by access to additional data from EirGrid, revealed that the PV installed capacity available on the spreadsheets is the Maximum Export Capacity (MEC) and does not accurately reflect the true generating capacity. Instead, MEC values from the spreadsheet are used as a basis. To adjust for actual grid conditions, installed capacities are assumed to be 1.4 times the MEC values. Whenever farm generation is calculated, this scaled capacity is used and any farm output is capped to the MEC.

@@

Atlite and C3S-E datasets are driven by the ERA5 reanalysis~\cite{hersbach2020era5}, produced by the European Centre for Medium-Range Weather Forecasts (ECMWF). This global gridded dataset provides hourly atmospheric variables from 1940 to the present. It is widely used for estimating PV and wind energy~\cite{mockert2023drought, dubus2023energy, brown2021drought, otero2022drought}. Table~\ref{tab:var_name} lists the ERA5 variables used in the models to generate the Atlite and C3S-Energy datasets.

@@

The C3S-E dataset estimates wind energy using wind speeds at 100 metres ($u_{100}$, $v_{100}$) and a standard turbine model, the Vestas V136/3450, with a fixed hub height of 100 m. This choice is based on expert advice and actual trend in wind turbine installation. The PV generation model used by C3S-E uses two ERA5 variables: global horizontal irradiance (referred to here as $ssrd$) and air temperature ($t2m$). PV generation is calculated multiple times using the same model with different azimuth and tilt angles. The results are aggregated based on a statistical distribution of the module angles based on the geographical location~\cite{saintdrenan2018solar}.

@@

Given that PV capacity in Ireland remains relatively low in 2023, and in order to explore how a more balanced distribution of wind and PV might impact RES droughts, this study also considers a balanced scenario, where the installed PV capacity is assumed to increase to 8 GW, while wind capacity rises to 9 GW. These figures are based on the target energy capacity outlined in the roadmap published by EirGrid~\cite{eirgrid2023future}. In this scenario, new PV time series are generated for both the Atlite and C3S-E G models, while the wind CF time series remain the same, as significant changes in wind farm distribution are not expected. From there, the results of these new time series are combined using updated weights which reflect a more balanced scenario (9 GW for wind and 8 GW for PV). \textcolor{magenta}{Is the specific distribution of the capacities mentioned somewhere? Same locations but scaled for wind, change in locations for PV}

@@

\begin{figure}[ht!]
	\centering
	\includegraphics[width=\textwidth]{droughts_methodology}
	\caption{Wind time series of CF (red) and its 24-hour moving average (green) from the 7th to the 15th of July 2021. The black dashed line indicates the CF threshold. The grey bar shows the period identified as a wind drought under our definition}
	\label{fig:find_res_droughts}
\end{figure}

@@

The C3S-E datasets use the Vestas V136/3450 wind turbine power curve, (Fig.~\ref{fig:power_curve}a). The Atlite model allows the user to specify the power curve. We considered the 121 power curves available for download from Renewables.ninja~\cite{staffell2016wake}. For each power curve, Renewables.ninja also provide four associated smoothed power curves. The smoothing is done by using a Gaussian filter with different standard deviations that depend on the wind speed. A wind CF time series for Ireland for all wind turbine power curve options and all smoothings is generated. The performance of each wind turbine power curve is then calculated based on four skill scores: correlation coefficient (CC), root mean square error (RMSE), mean bias error (MBE), and area under the curve. The area under the curve was calculated from histograms of the hourly CF values for the most recent decade, 2014-2023. Based on these metrics, the most representative power curve for Ireland was the Enercon E112.4500 power curve with the $0.3w$  smoothing filter.

@@

The smoothing of the wind turbine power curve represents losses due to the wake effect \textcolor{magenta}{yes, but not only. we do not want someone to be able to suck us into what the smoothing means rabbithole. My proposal: The smoothing of the wind turbine power curve represents losses associated with each individual turbine and the interaction between the different turbines}, which is important when modelling wind energy on larger scales. The histogram in Fig.~\ref{fig:power_curve}b shows that the C3S-E power curve tends to underestimate low CF values and overestimate higher ones, whereas the smoothed Atlite power curve more closely follows the recorded wind availability data from EirGrid.

@@

The effect of the differences in the power curves is also visible in Fig.~\ref{fig:wind_verification_contour} where the Atlite model shows good agreement with the EirGrid data. In contrast, the two C3S-E datasets tend to overestimate the EirGrid CF. This is confirmed by Table~\ref{tab:wind_skill_scores}, which shows that Atlite performs better than the C3S-E datasets for all three skill scores.

\textcolor{magenta}{I was reading Damian's comment about the tables being "a lot of numbers". I think the way it is written now is clear but please review it and see if you are happy with it too. I though about maybe explaining that RMSE and MBE are better closer to 0 and CC closer to 1, but honestly I think that the reader will know that, it is quite a common method and not a complex selection of skill scores.}

@@

The C3S-E datasets calculate PV generation multiple times using the same model with varying azimuth and tilt angles. The results are then aggregated based on a statistical distribution linked to each geographical location. Similarly to the wind CF verification, the Atlite model allows the user to select a given PV panel characteristics. In this study we have chosen to look at the three PV panels offered by in the Atlite model. Following the same methodology as in the previous section, we compare the three available models using four skill scores (CC, RMSE, MB, and area under the curve). Based on the best-performing metrics, we selected the Breyer PV panel model \cite{beyer2004pv}, using the Kaneka Hybrid panel option.

@@

Figure \ref{fig:solar_verification_contour} shows that the three datasets have a similar tendency of overestimating the CF compared to EirGrid, especially for high CF values. The skill scores presented in Table~\ref{tab:pv_skill_scores} indicate that C3S-E G performs best overall, with the lowest RMSE and a high correlation coefficient, suggesting a closer match to observed data. Atlite shows a slightly lower correlation and a higher RMSE, indicating larger deviations. All models show a slight positive bias, with Atlite exhibiting the most overestimation. These results reflect the broader challenges in accurately modeling PV, where variability is hard to capture. Nonetheless, C3S-E G’s lower RMSE suggests it may be more reliable for PV CF modeling in this context.

@@

Fig.~\ref{fig:bar_number_events_verification_pv} shows the number of PV drought events in 2023 across different duration ranges. The figure reveals partial agreement between the three datasets and the EirGrid data, with consistent results observed for duration ranges of 1-2, 3-4, 7-8, and 8+ days. However, discrepancies appear in the other ranges, where the models diverge from the EirGrid data. The main challenge in validating PV data stems from the recent installation of a large share of Ireland’s PV capacity, leading to uncertainties in PV generation data and the actual generating capacity in the first few months after each farm is connected. With over 60\% of the total PV capacity installed in 2023, these data uncertainties significantly impact the ability to perform rigorous validation for PV drought events. While the results are reasonably aligned in some ranges, the inherent challenges in PV data highlight the limitations of this verification.

@@

The median, first, and third quartiles for the Atlite dataset are consistently greater than or equal to those of the other two datasets, regardless of the duration range or type of renewable energy considered. This difference arises from the wind turbine power curve model used in the C3S-E datasets, which tends to overestimate the wind CF (Fig.~\ref{fig:wind_verification_contour}). As a result, the overall number of RES droughts is underestimated in the C3S-E datasets compared to Atlite. \textcolor{magenta}{Sorry to add this here again, but where are we on the current/projected/balanced thing? Just please make sure it is consistent, we can figure it out once Conor has gone over it, it's not crucial to the reading but I just remembered.}

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves significantly different than the two C3S-E datasets. The Atlite results follow the ones for wind with a log-linear increase, but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets likely arises from how each model handles PV output near the threshold. Atlite appears to hover slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions. \textcolor{magenta}{Not sure if it is worth a mention, but in case you had not made the connection: this is consistent with atlite having the highest MBE out of the three in the verification}

@@

Similar to previous results, the existing installed capacity wind-PV combination (Fig.\ref{fig:res_droughts_seasonality}c) shows patterns comparable to the wind panel (Fig.\ref{fig:res_droughts_seasonality}a). However, in this panel, the number of hours classified as RES droughts in summer decreases slightly compared to the wind-only scenario. This reduction can be explained by the contribution of PV generation during the summer months in the existing scenario, even though it constitutes only 11\% of total capacity. Since the number of RES drought hours for PV in summer is near zero, this small contribution has a noticeable impact on reducing overall drought hours. In the projected scenario (Fig.~\ref{fig:res_droughts_seasonality}d), all three datasets show a reduction in monthly RES drought frequency. The balanced mix of wind and PV in this scenario reduces the seasonal signal, slightly increasing the percentage of RES drought hours in winter months while significantly decreasing it in summer. Annual reductions in median RES drought frequency are observed across the datasets, dropping from 14\% to 5\% for Atlite, from 8\% to 3\% for C3S-E G, and from 9\% to 4\% for C3S-E N. \textcolor{magenta}{This is a good point to mention the seasonality of the demand in Ireland that Damian brought up, although it could also go in the conclusions}

@@

\textcolor{magenta}{So, here the panel names are different as in the other ones, with existing and balanced. this is what I meant by consistency. Also, note the panel a in the figure is labelled as c. that should be changed. }
